Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)

### 2023-11-20 manu-draft.qmd
update abstract from rgn comments

*Key words*: Bayesian, dissolved oxygen, ecosystem metabolism, estuary, gas exchange, open source

@@

Continuous monitoring data for dissolved oxygen have been used to estimate the contribution of gross primary production and ecosystem respiration to net ecosystem metabolism. Most methods for estimating metabolic processes make several assumptions on how biological and physical factors affect the expression of metabolism in dissolved oxygen data, which can lead to poorly constrained errors that influence the interpretation of results. In estuaries, these methods can inform our understanding of spatial and temporal patterns in trophic state, yet very few models have been developed that address challenges of estimating metabolism in these dynamic systems. This paper describes the development, assessment, and interpretation of results from a new approach for estimating metabolic rates in estuaries, the Estuarine BAyesian Single-station Estimation (EBASE) method. EBASE provides several advantages over existing models by using an integration formula appropriate for estuarine systems and using a Bayesian framework where critical model parameters are fit as a function of the data as informed by a set of prior distributions. This approach can also vary the time period over which critical model parameters are estimated. We demonstrate that EBASE can reproduce known metabolic parameters from a synthetic time series created with continuous monitoring data from the Apalachicola Bay (southeastern United States) and provide a quantitative assessment of how the prior distributions affect the model results. The approach is also bench-marked with more conventional methods for estimating metabolism, with critical differences attributed to formulation of the metabolic mass-balance equations and the underlying statistical framework (i.e., Bayesian vs simple integration). The EBASE approach is available using open source software and can be readily applied to multiple years of long-term monitoring data that are available in many estuaries. Overall, EBASE provides an accessible method to parameterize a simple metabolic model appropriate for estuarine systems and will provide additional understanding of processes that influence ecosystem status and condition.

*Key words*: Bayesian, dissolved oxygen, ecosystem metabolism, estuary, gas exchange, open source

@@

For ease of application, an R package was developed to implement EBASE [@Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>). EBASE differs from BASE in model implementation in two ways. First, users can specify the model optimization period, where the time period can vary from a minimum of 1 day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at 1 day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. Second, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters. Results obtained from interpolated data are automatically removed from the output, eliminating the need to pre- and post-process the data when using EBASE. Again, this feature was not included in the original BASE method and a complete time series was required for use. A detailed web page (<https://fawda123.github.io/EBASE/>) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the oxygen budget terms (P, R, and D), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters. All results provided herein were generated using the EBASE R package.

@@

As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of noise to the original synthetic time series. The objective of the analysis is to develop an expectation of how metabolic estimates may change using *in situ* data with noise that may substantially mask a biological signal in dissolved oxygen measurements. The noise estimates were created using a weighted regression analysis [@Beck15; @Beck21] on observed dissolved oxygen time series for Apalachicola Bay in 2021. This approach models dissolved oxygen as a function of time (day), hour, and water depth using a moving window approach with weights of the independent variables appropriate for each time period within the windows. The results of this analysis return a predicted and detided dissolved oxygen time series. Two separate time series of noise, representing residual (any other random process not related to metabolism or tidal advection) or tidal noise, were derived from the observed, predicted and detided time series. The residual noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series (Figure [-@fig-synapanos]a). Both the residual and tidal noise were added to the original synthetic time series in Figure [-@fig-synapa]c and are shown in Figure [-@fig-synapanos]b. EBASE was then used on the synthetic time series with noise, then compared with results from the original synthetic time series. The default priors justified in @sec-ebasemod (Equations [-@eq-aprior] and [-@eq-rprior]) were used with a 7 day optimization period. The prior for the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following similar logic used for the sensitivity analysis described above.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08. For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [as available in the WtRegDO R package, @Beck21]. An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood-based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic-based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected based on the alternative statistical approaches to estimate the metabolic results. Each comparison also evaluated differences in the results using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE used the default values justified in @sec-ebasemod (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). Similar to the previous analyses, the prior distribution for the $b$ parameter was fixed given the difficulty in recovery. A 1 day model optimization period was used for EBASE for comparability with the Odum method that returns daily estimates.

@@

As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE [@Grace15], which was developed for stream applications and some assumptions of the underlying model of the latter were inappropriate for the former. The most critical change made to BASE was adopting the gas transfer velocity parameterization in @Wanninkhof14, which is primarily wind-based and more appropriate than the depth-dependent formulation in BASE. EBASE also removed the temperature dependency of respiration, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS).

>>>>

*Key words*: Bayesian, dissolved oxygen, ecosystem metabolism, estuary, gas exchange, open-source

@@

Continuous measurements of dissolved oxygen (DO) are useful for quantifying ecosystem metabolism, which is critical for understanding estuarine biogeochemistry and ecology, but current methods applied to these data may lead to unphysical results and poorly constrained errors. Here we present a new approach for estimating estuarine metabolism: EBASE (Estuarine BAyesian Single-station Estimation). EBASE applies a Bayesian framework to a simple process-based model and DO observations, allowing the estimation of critical model parameters, specifically light efficiency and respiration, as informed by a set of prior distributions. EBASE improves upon the stream-based model from which it was derived by accommodating missing DO data and allowing the user to set the time period over which parameters are estimated. We demonstrate that EBASE can recover known metabolic parameters from a synthetic time series, even in the presence of noise (e.g., due to tidal advection) and when prior distributions are uninformed. Optimization periods of 7 days and 30 days are more preferable than 1 day. A comparison with the more-conventional method of Odum reveals the ability of EBASE to avoid unphysical results (such as negative photosynthesis and respiration) and improves when the DO data are detided. EBASE is available using open-source software and can be readily applied to multiple years of long-term monitoring data that are available in many estuaries. Overall, EBASE provides an accessible method to parameterize a simple metabolic model appropriate for estuarine systems and will provide additional understanding of processes that influence ecosystem status and condition.

@@

For ease of application, an R package was developed to implement EBASE [@Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>). EBASE differs from BASE in model implementation in two ways. First, users can specify the model optimization period, where the time period can vary from a minimum of 1 day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at 1 day. Sensitivity of the EBASE method to different optimization periods is described below. Second, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters. Results obtained from interpolated data are automatically removed from the output, eliminating the need to pre- and post-process the data when using EBASE. Again, this feature was not included in the original BASE method and a complete time series was required for use. A detailed web page (<https://fawda123.github.io/EBASE/>) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the oxygen budget terms (P, R, and D), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters. All results provided herein were generated using the EBASE R package.

@@

As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of noise to the original synthetic time series. The objective of the analysis is to develop an expectation of how metabolic estimates may change using *in situ* data with noise that may substantially mask a biological signal in dissolved oxygen measurements. The noise estimates were created using a weighted regression analysis [@Beck15; @Beck21] on observed dissolved oxygen time series for Apalachicola Bay in 2021. This approach models dissolved oxygen as a function of time (day), hour, and water depth using a moving window approach with weights of the independent variables appropriate for each time period within the windows. The results of this analysis return a predicted and detided dissolved oxygen time series. Two separate time series of noise, representing residual (any other random process not related to metabolism or tidal advection) or tidal noise, were derived from the observed, predicted and detided time series. The residual noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series (Figure [-@fig-synapanos]a). Both the residual and tidal noise were added to the original synthetic time series in Figure [-@fig-synapa]c and are shown in Figure [-@fig-synapanos]b. EBASE was then used on the synthetic time series with noise, then compared with results from the original synthetic time series. The default priors (Equations [-@eq-aprior] and [-@eq-rprior]) were used with a 7 day optimization period. The prior for the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following similar logic used for the sensitivity analysis described above.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08. For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [as available in the WtRegDO R package, @Beck21]. An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood-based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic-based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected based on the alternative statistical approaches to estimate the metabolic results. Each comparison also evaluated differences in the results using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE used the default values (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). Similar to the previous analyses, the prior distribution for the $b$ parameter was fixed given the difficulty in recovery. A 1 day model optimization period was used for EBASE for comparability with the Odum method that returns daily estimates.

@@

As described in the materials and procedures section, EBASE is an estuarine adaptation of BASE [@Grace15], which was developed for stream applications and some assumptions of the underlying model of the latter were inappropriate for the former. The most critical change made to BASE was adopting the gas transfer velocity parameterization in @Wanninkhof14, which is primarily wind-based and more appropriate than the depth-dependent formulation in BASE. EBASE also removed the temperature dependency of respiration, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS).
